---
title: 'Rcode: COMPSAC 2024'
author: "Sulalitha Bowala"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list=ls(all=TRUE)) # Remove objects from environment
set.seed(12)
```



```{r, message=FALSE}
### Call packages
require("tseries") # load the tseries library
require("zoo")
require("moments")
require("car")
require("MASS")
require("stats")  
require("VGAM")
require("fGarch")
require("quantmod") # To download data alternatively
require("PerformanceAnalytics") # To draw Timeseries graphs
require("scales")
require("PEIP")
require("dplyr")
require("lubridate")
require("tidyverse")
require("gridExtra")
require("gdata")
require("xtable")
require("vioplot")
require("fpp3")
require("readr")
require("tis")
require("tibble")
require("tsibble")
require("stringr")
require("magrittr")
require("fable")
require("Metrics") # rmse()
require("forecast")
```



```{r}
## Import data
## Ontario data

Ontario_demand_hourly <- read.csv("Data_2018_2019.csv")
```


```{r}
### Manipulate hourly data to obtain daily data
### + create a new categorical variable (day-type)

Ontario_demand_hourly$Time <- as.POSIXct(Ontario_demand_hourly$Time, "%Y-%m-%d %H:%M:%S")
Ontario_demand_hourly$Temperature <- Ontario_demand_hourly$Temperature
Ontario_demand_hourly$Precipitation <- Ontario_demand_hourly$Precipitation
Ontario_demand_hourly$CloudCover <- Ontario_demand_hourly$CloudCover
Ontario_demand_hourly$IrradianceSurface <- Ontario_demand_hourly$IrradianceSurface
Ontario_demand_hourly$IrradianceToa <- Ontario_demand_hourly$IrradianceToa
Ontario_demand_hourly$Snowfall <- Ontario_demand_hourly$Snowfall
Ontario_demand_hourly$SnowMass <- Ontario_demand_hourly$SnowMass
Ontario_demand_hourly$Location <- Ontario_demand_hourly$Location
Ontario_demand_hourly <- as_tsibble(Ontario_demand_hourly,index = 'Time',key = 'Location')

Ontario_demand_hourly <- Ontario_demand_hourly %>% 
  mutate(
    Date = date( Time)
  )
Ontario_demand_hourly <- Ontario_demand_hourly %>% 
  mutate(
    Holiday = isHoliday(Ontario_demand_hourly$Date)
  )
Ontario_demand_hourly <- Ontario_demand_hourly %>%
 mutate(Day_Type = case_when(
    Holiday ~ "Holiday",
    wday(Date) %in% 2:6 ~ "Weekday",
    TRUE ~ "Weekend"
  ))
# 
head(Ontario_demand_hourly)
tail(Ontario_demand_hourly)
```



```{r}
## Rename columns

Ontario_demand_hourly<-Ontario_demand_hourly[,c("Time","Precipitation","Temperature","IrradianceSurface","IrradianceToa","Snowfall","SnowMass","CloudCover","AirDensity","Demand","Location","Date","Holiday","Day_Type")]
```




```{r}
## Daily data

# Ontario data
Ontario.Daily <- Ontario_demand_hourly %>%

  index_by(Date = date(Time)) %>%
  summarise(
    Demand = mean(Demand),
    Temperature = mean(Temperature),
    Holiday = any(Holiday),
    Precipitation = mean(Precipitation),
    IrradianceSurface = mean(IrradianceSurface),
    IrradianceToa = mean(IrradianceToa),
    Snowfall = mean(Snowfall),
    SnowMass = mean(SnowMass),
    CloudCover = mean(CloudCover),
    AirDensity = mean(AirDensity),
  ) %>%
  mutate(Day_Type = case_when(
    Holiday ~ "Holiday",
    wday(Date) %in% 2:6 ~ "Weekday",
    TRUE ~ "Weekend"
  ))

Ontario.Daily <- Ontario.Daily %>%
  mutate(Temperature2 = Temperature^2,
         HolidayNew = if_else(Holiday == "TRUE", 1, 0),
         Day_TypeNew = if_else(Day_Type == "Weekday", 1, 0))

head(Ontario.Daily)
tail(Ontario.Daily)
```


whether data: https://www.renewables.ninja/

We believe Temperature is affected by other covariates (Precipitation, IrradianceSurface, IrradianceToa, Snowfall, SnowMass, CloudCover, AirDensity)  available in the data.

Description of covariates:

Air temperature (°C): Air temperature at 2 meters above ground. T2M variable in MERRA-2, converted from K.

Precipitation (mm / hour): Total bias-corrected precipitation, over land only. PRECTOTLAND variable in MERRA-2, converted from kg m⁻² s⁻¹

Snowfall (mm / hour): Total bias-corrected precipitation in the form of snow, over land only. PRECSNOLAND variable in MERRA-2, converted from kg m⁻² s⁻¹

Snow mass (kg / m²): Amount of snow per land area. SNOMAS variable in MERRA-2, in native units.

Air density (kg / m³): Air density at ground level. RHOA variable in MERRA-2, in native units.

Ground-level solar irradiance (W / m²): Surface-level incident shortwave radiation flux, considering cloud cover and aerosols. SWGDN variable in MERRA-2, in native units. Note, MERRA-2 uses the mean annual climate for aerosols, rather than time-varying quantity.

Top of atmosphere solar irradiance (W / m²): Top of atmosphere incident shortwave radiation flux, before cloud cover and aerosol influences. SWTDN variable in MERRA-2, in native units.

Cloud cover fraction: Fraction of cloud cover, averaged over grid cell and summed over all height above ground. CLDTOT variable in MERRA-2, in native units (a [0, 1] scale).


Check for multicollinearity among selected variables (Precipitation, IrradianceSurface, IrradianceToa, Snowfall, SnowMass, CloudCover, AirDensity).


```{r}
## VIF (Variance Inflation Factor)

library(car)

model.VIF <- lm(Temperature ~ Precipitation + IrradianceSurface + IrradianceToa + Snowfall + SnowMass + CloudCover + AirDensity, data = Ontario.Daily)

summary(model.VIF)

# Calculate VIF for selected variables
vif_values <- vif(model.VIF)

# Print VIF values
print(vif_values)
```


The Variance Inflation Factor (VIF) is a measure that quantifies how much the variance of an estimated regression coefficient increases if your predictors are correlated. High VIF values (typically > 10) may indicate problematic levels of multicollinearity.

Only for IrradianceSurface we observe a VIF value greater than 10 and VIF of IrradianceToa is close to 10.


```{r}
## Correlation Matrix

Temp.Ontario.Daily <- Ontario.Daily

colnames(Temp.Ontario.Daily) <- c("Date","Demand","Temperature","Holiday",
                                  "Precipitation","Irradiance (surface)",
                                  "Irradiance (atmosphere)","Snowfall",
                                  "Snow Mass","Cloud Cover","Air Density",
                                  "Day_Type","Temperature2",
                                  "HolidayNew","Day_TypeNew")

# Create a correlation matrix for selected variables
cor_matrix <- cor(Temp.Ontario.Daily[, c("Precipitation", "Irradiance (surface)", "Irradiance (atmosphere)", "Snowfall",  "Snow Mass",  "Cloud Cover", "Air Density")])

# Print the correlation matrix
print(cor_matrix)
```


High correlation coefficients (close to 1 or -1) may suggest multicollinearity. 

Corr(IrradianceToa, IrradianceSurface) = 0.8585024
Corr(IrradianceSurface, CloudCover) = -0.6143980
Corr(IrradianceSurface, AirDensity) = -0.6453389
Corr(IrradianceToa, AirDensity) = -0.7476595


We can visualize the correlation matrix using tools like corrplot for better insight:

```{r}
library(corrplot)

# Plot the correlation matrix
corrplot(cor_matrix, method = "color", tl.col="black")
```


Thus, we may omit "IrradianceSurface" and "IrradianceToa" from future models for Temperature to avoid multicollinearity.


## Temperature forecasts

```{r}
## Training (75%) and testing (25%) datasets

Temp.Train <- Ontario.Daily %>% filter(Date <= "2019-06-30")
Temp.Test <- Ontario.Daily %>% filter(Date > "2019-06-30")
```



```{r}
library(fable.prophet)
## Prophet Model (with covariates)
Prophet.Model.Temp2 <- Temp.Train %>%
  model(prophet(Temperature ~ Precipitation + Snowfall + SnowMass + 
                  CloudCover + AirDensity))
fore.Prophet.Temp2 <- Prophet.Model.Temp2 %>% forecast(Temp.Test)
Prophet.Test.Accuracy2 <- fabletools::accuracy(fore.Prophet.Temp2, Temp.Test)
xtable(Prophet.Test.Accuracy2)
```



```{r}
## ARIMA Model (with covariates)
ARIMA.Model.Temp2 <- Temp.Train %>%
  model(ARIMA(Temperature ~ Precipitation + Snowfall + SnowMass + CloudCover + AirDensity))
ARIMA.Model.Temp2
fore.ARIMA.Temp2 <- ARIMA.Model.Temp2 %>% forecast(Temp.Test)
ARIMA.Test.Accuracy2 <- fabletools::accuracy(fore.ARIMA.Temp2, Temp.Test)
xtable(ARIMA.Test.Accuracy2)

```



```{r}
## nnetar Model (with covariates)
nnetar.Model.Temp2 <- Temp.Train %>%
  model(NNETAR(Temperature ~ Precipitation + Snowfall + SnowMass + CloudCover + AirDensity))
nnetar.Model.Temp2
fore.nnetar.Temp2 <- nnetar.Model.Temp2 %>% forecast(Temp.Test)
nnetar.Test.Accuracy2 <- fabletools::accuracy(fore.nnetar.Temp2, Temp.Test)
xtable(nnetar.Test.Accuracy2)
```



```{r}
## TSLM (with covariates)
TSLM.Model.Temp2 <- Temp.Train %>%
  model(TSLM(Temperature ~ Precipitation + Snowfall + SnowMass + CloudCover + AirDensity))
fore.TSLM.Temp2 <- TSLM.Model.Temp2 %>% forecast(Temp.Test)
TSLM.Temp.Accuracy2 <- fabletools::accuracy(fore.TSLM.Temp2, Temp.Test)
xtable(TSLM.Temp.Accuracy2)

```


