---
title: 'Rcode: COMPSAC 2024'
author: "Sulalitha Bowala"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list=ls(all=TRUE)) # Remove objects from environment
set.seed(12)
```



```{r, message=FALSE}
### Call packages
require("tseries") # load the tseries library
require("zoo")
require("moments")
require("car")
require("MASS")
require("stats")  
require("VGAM")
require("fGarch")
require("quantmod") # To download data alternatively
require("PerformanceAnalytics") # To draw Timeseries graphs
require("scales")
require("PEIP")
require("dplyr")
require("lubridate")
require("tidyverse")
require("gridExtra")
require("gdata")
require("xtable")
require("vioplot")
require("fpp3")
require("readr")
require("tis")
require("tibble")
require("tsibble")
require("stringr")
require("magrittr")
require("fable")
require("Metrics") # rmse()
require("forecast")
```



```{r}
# Download data
start.date = '2016-01-01' # starting date of stock
end.date = '2018-12-31' # ending date of stock

# Download Google (GOOG) from Yahoo finance
getSymbols("GOOG", src = "yahoo", from = start.date, to = end.date)
GOOG <- na.omit(GOOG)
# create date variable
GOOG <- zoo::fortify.zoo(GOOG)
# Rename date variable
GOOG <- GOOG %>% rename("Date" = "Index")
# create tissble object
GOOG <- as_tsibble(GOOG, index = Date) # create tissble object
# Re-index based on trading days (as there are some missing days)
GOOG_stock <- GOOG |>
  mutate(day = row_number()) |>
  update_tsibble(index = day, regular = TRUE)



# Download VIX (^VIX) from Yahoo finance
getSymbols("^VIX", src = "yahoo", from = start.date, to = end.date)
VIX <- na.omit(VIX)
# create date variable
VIX <- zoo::fortify.zoo(VIX)
# Rename date variable
VIX <- VIX %>% rename("Date" = "Index")
# create tissble object
VIX <- as_tsibble(VIX, index = Date) # create tissble object
# Re-index based on trading days (as there are some missing days)
VIX_stock <- VIX |>
  mutate(day = row_number()) |>
  update_tsibble(index = day, regular = TRUE)


# Download Bitcoin (BTC-USD) from Yahoo finance
BTC <- getSymbols("BTC-USD", from = start.date, to = end.date, source="yahoo", auto.assign=FALSE, return.class="xts")
BTC <- na.omit(BTC)
# create date variable
BTC <- zoo::fortify.zoo(BTC)
# Rename date variable
BTC <- BTC %>% rename("Date" = "Index")
# create tissble object
BTC <- as_tsibble(BTC, index = Date) # create tissble object
# Re-index based on trading days (as there are some missing days)
BTC_stock <- BTC |>
  mutate(day = row_number()) |>
  update_tsibble(index = day, regular = TRUE)
colnames(BTC_stock) <- c("Date","BTC.Open","BTC.High","BTC.Low","BTC.Close","BTC.Volume","BTC.Adjusted","day")

```



```{r}
## Import data
## Ontario data

Ontario_demand_hourly <- read.csv("Data_2018_2019.csv")
```


```{r}
### Manipulate hourly data to obtain daily data
### + create a new categorical variable (day-type)

Ontario_demand_hourly$Time <- as.POSIXct(Ontario_demand_hourly$Time, "%Y-%m-%d %H:%M:%S")
Ontario_demand_hourly$Temperature <- Ontario_demand_hourly$Temperature
Ontario_demand_hourly$Precipitation <- Ontario_demand_hourly$Precipitation
Ontario_demand_hourly$CloudCover <- Ontario_demand_hourly$CloudCover
Ontario_demand_hourly$IrradianceSurface <- Ontario_demand_hourly$IrradianceSurface
Ontario_demand_hourly$IrradianceToa <- Ontario_demand_hourly$IrradianceToa
Ontario_demand_hourly$Snowfall <- Ontario_demand_hourly$Snowfall
Ontario_demand_hourly$SnowMass <- Ontario_demand_hourly$SnowMass
Ontario_demand_hourly$Location <- Ontario_demand_hourly$Location
Ontario_demand_hourly <- as_tsibble(Ontario_demand_hourly,index = 'Time',key = 'Location')

Ontario_demand_hourly <- Ontario_demand_hourly %>% 
  mutate(
    Date = date( Time)
  )
Ontario_demand_hourly <- Ontario_demand_hourly %>% 
  mutate(
    Holiday = isHoliday(Ontario_demand_hourly$Date)
  )
Ontario_demand_hourly <- Ontario_demand_hourly %>%
 mutate(Day_Type = case_when(
    Holiday ~ "Holiday",
    wday(Date) %in% 2:6 ~ "Weekday",
    TRUE ~ "Weekend"
  ))
# 
head(Ontario_demand_hourly)
tail(Ontario_demand_hourly)
```


```{r}
## Rename columns

Ontario_demand_hourly<-Ontario_demand_hourly[,c("Time","Precipitation","Temperature","IrradianceSurface","IrradianceToa","Snowfall","SnowMass","CloudCover","AirDensity","Demand","Location","Date","Holiday","Day_Type")]
```




```{r}
## Daily data

# Ontario data
Ontario.Daily <- Ontario_demand_hourly %>%

  index_by(Date = date(Time)) %>%
  summarise(
    Demand = mean(Demand),
    Temperature = mean(Temperature),
    Holiday = any(Holiday),
    Precipitation = mean(Precipitation),
    IrradianceSurface = mean(IrradianceSurface),
    IrradianceToa = mean(IrradianceToa),
    Snowfall = mean(Snowfall),
    SnowMass = mean(SnowMass),
    CloudCover = mean(CloudCover),
    AirDensity = mean(AirDensity),
  ) %>%
  mutate(Day_Type = case_when(
    Holiday ~ "Holiday",
    wday(Date) %in% 2:6 ~ "Weekday",
    TRUE ~ "Weekend"
  ))

Ontario.Daily <- Ontario.Daily %>%
  mutate(Temperature2 = Temperature^2,
         HolidayNew = if_else(Holiday == "TRUE", 1, 0),
         Day_TypeNew = if_else(Day_Type == "Weekday", 1, 0))

head(Ontario.Daily)
tail(Ontario.Daily)
```


whether data: https://www.renewables.ninja/

We believe Temperature is affected by other covariates (Precipitation, IrradianceSurface, IrradianceToa, Snowfall, SnowMass, CloudCover, AirDensity)  available in the data.

Description of covariates:

Air temperature (°C): Air temperature at 2 meters above ground. T2M variable in MERRA-2, converted from K.

Precipitation (mm / hour): Total bias-corrected precipitation, over land only. PRECTOTLAND variable in MERRA-2, converted from kg m⁻² s⁻¹

Snowfall (mm / hour): Total bias-corrected precipitation in the form of snow, over land only. PRECSNOLAND variable in MERRA-2, converted from kg m⁻² s⁻¹

Snow mass (kg / m²): Amount of snow per land area. SNOMAS variable in MERRA-2, in native units.

Air density (kg / m³): Air density at ground level. RHOA variable in MERRA-2, in native units.

Ground-level solar irradiance (W / m²): Surface-level incident shortwave radiation flux, considering cloud cover and aerosols. SWGDN variable in MERRA-2, in native units. Note, MERRA-2 uses the mean annual climate for aerosols, rather than time-varying quantity.

Top of atmosphere solar irradiance (W / m²): Top of atmosphere incident shortwave radiation flux, before cloud cover and aerosol influences. SWTDN variable in MERRA-2, in native units.

Cloud cover fraction: Fraction of cloud cover, averaged over grid cell and summed over all height above ground. CLDTOT variable in MERRA-2, in native units (a [0, 1] scale).


Check for multicollinearity among selected variables (Precipitation, IrradianceSurface, IrradianceToa, Snowfall, SnowMass, CloudCover, AirDensity).


```{r}
## VIF (Variance Inflation Factor)

library(car)

model.VIF <- lm(Temperature ~ Precipitation + IrradianceSurface + IrradianceToa + Snowfall + SnowMass + CloudCover + AirDensity, data = Ontario.Daily)

summary(model.VIF)

# Calculate VIF for selected variables
vif_values <- vif(model.VIF)

# Print VIF values
print(vif_values)
```


The Variance Inflation Factor (VIF) is a measure that quantifies how much the variance of an estimated regression coefficient increases if your predictors are correlated. High VIF values (typically > 10) may indicate problematic levels of multicollinearity.

Only for IrradianceSurface we observe a VIF value greater than 10 and VIF of IrradianceToa is close to 10.


```{r}
## Correlation Matrix

Temp.Ontario.Daily <- Ontario.Daily

colnames(Temp.Ontario.Daily) <- c("Date","Demand","Temperature","Holiday",
                                  "Precipitation","Irradiance (surface)",
                                  "Irradiance (atmosphere)","Snowfall",
                                  "Snow Mass","Cloud Cover","Air Density",
                                  "Day_Type","Temperature2",
                                  "HolidayNew","Day_TypeNew")

# Create a correlation matrix for selected variables
cor_matrix <- cor(Temp.Ontario.Daily[, c("Precipitation", "Irradiance (surface)", "Irradiance (atmosphere)", "Snowfall",  "Snow Mass",  "Cloud Cover", "Air Density")])

# Print the correlation matrix
print(cor_matrix)
```


High correlation coefficients (close to 1 or -1) may suggest multicollinearity. 

Corr(IrradianceToa, IrradianceSurface) = 0.8585024
Corr(IrradianceSurface, CloudCover) = -0.6143980
Corr(IrradianceSurface, AirDensity) = -0.6453389
Corr(IrradianceToa, AirDensity) = -0.7476595


We can visualize the correlation matrix using tools like corrplot for better insight:

```{r}
library(corrplot)

# Plot the correlation matrix
corrplot(cor_matrix, method = "color", tl.col="black")
```


Thus, we may omit "IrradianceSurface" and "IrradianceToa" from future models for Temperature to avoid multicollinearity.



## Price forecasts (Google, VIX, BTC) and Temperature forecasts

```{r}
GOOG.Train <- GOOG_stock

VIX.Train <- VIX_stock

BTC.Train <- BTC_stock

Temp.Train <- Ontario.Daily

```


```{r}
plot(GOOG.Train$GOOG.Adjusted, type = "l")

plot(VIX.Train$VIX.Adjusted, type = "l")

plot(BTC.Train$BTC.Adjusted, type = "l")
```



## Price forecasts (Google)

```{r}
# Time series cross-validation accuracy (Creating Training dataset)
GOOG.Train.tr <- GOOG.Train |>
  stretch_tsibble(.init = 150, .step = 1) |>
  relocate(Date, .id)
head(GOOG.Train.tr)


# TSCV accuracy

## Naive model
Naive.Model.GOOG <- GOOG.Train.tr %>%
  model(NAIVE(GOOG.Adjusted))
fore.Naive.GOOG <- Naive.Model.GOOG %>% forecast(h=1)
Naive.GOOG.Accuracy <- fabletools::accuracy(fore.Naive.GOOG, GOOG.Train)
xtable(Naive.GOOG.Accuracy, digits = 4)


## Drift model
Drift.Model.GOOG <- GOOG.Train.tr %>%
  model(RW(GOOG.Adjusted~drift()))
fore.Drift.GOOG <- Drift.Model.GOOG %>% forecast(h=1)
Drift.GOOG.Accuracy <- fabletools::accuracy(fore.Drift.GOOG, GOOG.Train)
xtable(Drift.GOOG.Accuracy)


## Prophet model
library(fable.prophet)
Prophet.Model.GOOG <- GOOG.Train.tr %>%
  model(prophet(GOOG.Adjusted))
fore.Prophet.GOOG <- Prophet.Model.GOOG %>% forecast(h=1)
Prophet.GOOG.Accuracy <- fabletools::accuracy(fore.Prophet.GOOG, GOOG.Train)
xtable(Prophet.GOOG.Accuracy)


## ARIMA model
ARIMA.Model.GOOG <- GOOG.Train.tr %>%
  model(ARIMA(GOOG.Adjusted))
#ARIMA.Model.GOOG
fore.ARIMA.GOOG <- ARIMA.Model.GOOG %>% forecast(h=1)
ARIMA.GOOG.Accuracy <- fabletools::accuracy(fore.ARIMA.GOOG, GOOG.Train)
xtable(ARIMA.GOOG.Accuracy)
## ARIMA model for the study period
GOOG.Train %>% model(ARIMA(GOOG.Adjusted))


## nnetar model
nnetar.Model.GOOG <- GOOG.Train.tr %>%
  model(NNETAR(GOOG.Adjusted))
#nnetar.Model.GOOG
fore.nnetar.GOOG <- nnetar.Model.GOOG %>% forecast(h=1)
nnetar.GOOG.Accuracy <- fabletools::accuracy(fore.nnetar.GOOG, GOOG.Train)
xtable(nnetar.GOOG.Accuracy)
## nnetar model for the study period
GOOG.Train %>% model(NNETAR(GOOG.Adjusted))


## EWMA model
EWMA.Model.GOOG <- GOOG.Train.tr %>%
  model(ETS(GOOG.Adjusted))
fore.EWMA.GOOG <- EWMA.Model.GOOG %>% forecast(h=1)
EWMA.GOOG.Accuracy <- fabletools::accuracy(fore.EWMA.GOOG, GOOG.Train)
xtable(EWMA.GOOG.Accuracy)

```



## Price forecasts (VIX)

```{r}
# Time series cross-validation accuracy (Creating Training dataset)
VIX.Train.tr <- VIX.Train |>
  stretch_tsibble(.init = 150, .step = 1) |>
  relocate(Date, .id)
head(VIX.Train.tr)


# TSCV accuracy

## Naive model
Naive.Model.VIX <- VIX.Train.tr %>%
  model(NAIVE(VIX.Adjusted))
fore.Naive.VIX <- Naive.Model.VIX %>% forecast(h=1)
Naive.VIX.Accuracy <- fabletools::accuracy(fore.Naive.VIX, VIX.Train)
xtable(Naive.VIX.Accuracy)


## Drift model
Drift.Model.VIX <- VIX.Train.tr %>%
  model(RW(VIX.Adjusted~drift()))
fore.Drift.VIX <- Drift.Model.VIX %>% forecast(h=1)
Drift.VIX.Accuracy <- fabletools::accuracy(fore.Drift.VIX, VIX.Train)
xtable(Drift.VIX.Accuracy)


## Prophet model
#library(fable.prophet)
Prophet.Model.VIX <- VIX.Train.tr %>%
  model(prophet(VIX.Adjusted))
fore.Prophet.VIX <- Prophet.Model.VIX %>% forecast(h=1)
Prophet.VIX.Accuracy <- fabletools::accuracy(fore.Prophet.VIX, VIX.Train)
xtable(Prophet.VIX.Accuracy)


## ARIMA model
ARIMA.Model.VIX <- VIX.Train.tr %>%
  model(ARIMA(VIX.Adjusted))
#ARIMA.Model.VIX
fore.ARIMA.VIX <- ARIMA.Model.VIX %>% forecast(h=1)
ARIMA.VIX.Accuracy <- fabletools::accuracy(fore.ARIMA.VIX, VIX.Train)
xtable(ARIMA.VIX.Accuracy)
## ARIMA model for the study period
VIX.Train %>% model(ARIMA(VIX.Adjusted))


## nnetar model
nnetar.Model.VIX <- VIX.Train.tr %>%
  model(NNETAR(VIX.Adjusted))
#nnetar.Model.VIX
fore.nnetar.VIX <- nnetar.Model.VIX %>% forecast(h=1)
nnetar.VIX.Accuracy <- fabletools::accuracy(fore.nnetar.VIX, VIX.Train)
xtable(nnetar.VIX.Accuracy)
## nnetar model for the study period
VIX.Train %>% model(NNETAR(VIX.Adjusted))


## EWMA model
EWMA.Model.VIX <- VIX.Train.tr %>%
  model(ETS(VIX.Adjusted))
fore.EWMA.VIX <- EWMA.Model.VIX %>% forecast(h=1)
EWMA.VIX.Accuracy <- fabletools::accuracy(fore.EWMA.VIX, VIX.Train)
xtable(EWMA.VIX.Accuracy)

```



## Price forecasts (BTC)

```{r}
# Time series cross-validation accuracy (Creating Training dataset)
BTC.Train.tr <- BTC.Train |>
  stretch_tsibble(.init = 220, .step = 1) |>
  relocate(Date, .id)
head(BTC.Train.tr)


# TSCV accuracy

## Naive model
Naive.Model.BTC <- BTC.Train.tr %>%
  model(NAIVE(BTC.Adjusted))
fore.Naive.BTC <- Naive.Model.BTC %>% forecast(h=1)
Naive.BTC.Accuracy <- fabletools::accuracy(fore.Naive.BTC, BTC.Train)
xtable(Naive.BTC.Accuracy)


## Drift model
Drift.Model.BTC <- BTC.Train.tr %>%
  model(RW(BTC.Adjusted~drift()))
fore.Drift.BTC <- Drift.Model.BTC %>% forecast(h=1)
Drift.BTC.Accuracy <- fabletools::accuracy(fore.Drift.BTC, BTC.Train)
xtable(Drift.BTC.Accuracy)


## Prophet model
#library(fable.prophet)
Prophet.Model.BTC <- BTC.Train.tr %>%
  model(prophet(BTC.Adjusted))
fore.Prophet.BTC <- Prophet.Model.BTC %>% forecast(h=1)
Prophet.BTC.Accuracy <- fabletools::accuracy(fore.Prophet.BTC, BTC.Train)
xtable(Prophet.BTC.Accuracy)


## ARIMA model
ARIMA.Model.BTC <- BTC.Train.tr %>%
  model(ARIMA(BTC.Adjusted))
#ARIMA.Model.BTC
fore.ARIMA.BTC <- ARIMA.Model.BTC %>% forecast(h=1)
ARIMA.BTC.Accuracy <- fabletools::accuracy(fore.ARIMA.BTC, BTC.Train)
xtable(ARIMA.BTC.Accuracy)
## ARIMA model for the study period
BTC.Train %>% model(ARIMA(BTC.Adjusted))


## nnetar model
nnetar.Model.BTC <- BTC.Train.tr %>%
  model(NNETAR(BTC.Adjusted))
#nnetar.Model.BTC
fore.nnetar.BTC <- nnetar.Model.BTC %>% forecast(h=1)
nnetar.BTC.Accuracy <- fabletools::accuracy(fore.nnetar.BTC, BTC.Train)
xtable(nnetar.BTC.Accuracy)
## nnetar model for the study period
BTC.Train %>% model(NNETAR(BTC.Adjusted))


## EWMA model
EWMA.Model.BTC <- BTC.Train.tr %>%
  model(ETS(BTC.Adjusted))
fore.EWMA.BTC <- EWMA.Model.BTC %>% forecast(h=1)
EWMA.BTC.Accuracy <- fabletools::accuracy(fore.EWMA.BTC, BTC.Train)
xtable(EWMA.BTC.Accuracy)

```



## Temperature forecasts

```{r}
# Time series cross-validation accuracy (Creating Training dataset)
Temp.Train.tr <- Temp.Train |>
  stretch_tsibble(.init = 150, .step = 1) |>
  relocate(Date, .id)
head(Temp.Train.tr)
```



```{r}
# TSCV accuracy

## Mean model
Mean.Model.Temp <- Temp.Train.tr %>%
  model(MEAN(Temperature))
fore.Mean.Temp <- Mean.Model.Temp %>% forecast(h=1)
Mean.Temp.Accuracy <- fabletools::accuracy(fore.Mean.Temp, Temp.Train)
xtable(Mean.Temp.Accuracy)
```



```{r}
## Naive model
Naive.Model.Temp <- Temp.Train.tr %>%
  model(NAIVE(Temperature))
fore.Naive.Temp <- Naive.Model.Temp %>% forecast(h=1)
Naive.Temp.Accuracy <- fabletools::accuracy(fore.Naive.Temp, Temp.Train)
xtable(Naive.Temp.Accuracy)
```



```{r}
## Seasonal Naive model
SNaive.Model.Temp <- Temp.Train.tr %>%
  model(SNAIVE(Temperature))
fore.SNaive.Temp <- SNaive.Model.Temp %>% forecast(h=1)
SNaive.Temp.Accuracy <- fabletools::accuracy(fore.SNaive.Temp, Temp.Train)
xtable(SNaive.Temp.Accuracy)
```



```{r}
## Drift model
Drift.Model.Temp <- Temp.Train.tr %>%
  model(RW(Temperature~drift()))
fore.Drift.Temp <- Drift.Model.Temp %>% forecast(h=1)
Drift.Temp.Accuracy <- fabletools::accuracy(fore.Drift.Temp, Temp.Train)
xtable(Drift.Temp.Accuracy)
```



```{r}
## Prophet Model (without covariates)
Prophet.Model.Temp1 <- Temp.Train.tr %>%
  model(prophet(Temperature))
fore.Prophet.Temp1 <- Prophet.Model.Temp1 %>% forecast(h=1)
Prophet.Test.Accuracy1 <- fabletools::accuracy(fore.Prophet.Temp1, Temp.Train)
xtable(Prophet.Test.Accuracy1)
```



```{r}
## ARIMA Model (without covariates)
ARIMA.Model.Temp1 <- Temp.Train.tr %>%
  model(ARIMA(Temperature))
#ARIMA.Model.Temp1
fore.ARIMA.Temp1 <- ARIMA.Model.Temp1 %>% forecast(h=1)
ARIMA.Test.Accuracy1 <- fabletools::accuracy(fore.ARIMA.Temp1, Temp.Train)
xtable(ARIMA.Test.Accuracy1)
## ARIMA model for the study period
Temp.Train %>% model(ARIMA(Temperature))
```



```{r}
## nnetar Model (without covariates)
nnetar.Model.Temp1 <- Temp.Train.tr %>%
  model(NNETAR(Temperature))
nnetar.Model.Temp1
fore.nnetar.Temp1 <- nnetar.Model.Temp1 %>% forecast(h=1)
nnetar.Test.Accuracy1 <- fabletools::accuracy(fore.nnetar.Temp1, Temp.Train)
xtable(nnetar.Test.Accuracy1)
## nnetar model for the study period
Temp.Train %>% model(NNETAR(Temperature))
```



```{r}
## EWMA model
EWMA.Model.Temp <- Temp.Train.tr %>%
  model(ETS(Temperature))
fore.EWMA.Temp <- EWMA.Model.Temp %>% forecast(h=1)
EWMA.Temp.Accuracy <- fabletools::accuracy(fore.EWMA.Temp, Temp.Train)
xtable(EWMA.Temp.Accuracy)
```



```{r}
## TSLM (without covariates)
TSLM.Model.Temp1 <- Temp.Train.tr %>%
  model(TSLM(Temperature))
fore.TSLM.Temp1 <- TSLM.Model.Temp1 %>% forecast(h=1)
TSLM.Temp.Accuracy1 <- fabletools::accuracy(fore.TSLM.Temp1, Temp.Train)
xtable(TSLM.Temp.Accuracy1)
```



